% -*-latexinfo-*-
% plplot.tex
% Geoffrey Furnish
% 9 May 1994
%
% This is a LaTeXinfo document detailing the use of PLplot.
%
% Process via:	% latex plplot
%
% $Id$
% $Log$
% Revision 1.7  1994/10/11 19:00:19  furnish
% Primarily structuring corrections, some typos.  Nothing really major.
%
% Revision 1.6  1994/05/26  22:37:43  mjl
% Added disclaimers about the messy state of the document.
%
% Revision 1.5  1994/05/23  22:04:52  mjl
% Changed all \noinput's back to \input's.
%
% Revision 1.4  1994/05/19  22:44:50  mjl
% Changed to inputting the api section.
%
% Revision 1.3  1994/05/19  06:43:07  mjl
% Changed \documentstyle to include the plplot style file.  Not all
% sections included now; just the ones I'm working on (will need
% to be changed back).  Various fixes, updates, flashing lights...
%
% Revision 1.2  1994/05/09  22:09:06  furnish
% Improvements to the new doc.
%
% Revision 1.1  1994/05/09  15:06:25  furnish
% New files for building the PLplot manual via LaTeXinfo.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentstyle[12pt,fullpage,latexinfo,plplot]{book}
\pagestyle{headings}

\begin{document}

\c Declare which indices you want to make use of.
\newindex{cp}

\title{ The PLplot Plotting Library \\ 
        Programmer's Reference Manual \\
        Version 5.0}
\author{
        Maurice J. LeBrun\\
        Geoff Furnish\\
\\
        Institute for Fusion Studies\\
        University of Texas at Austin\\
        }

\date{\today}
\maketitle

\c The following commands start the copyright page for the printed manual.
\clearpage
\vspace{0pt plus 1filll}
Copyright \copyright\ 1994 Geoffrey Furnish, Maurice LeBrun.

PLplot may be used by anyone.

This manual may be copied by anyone.

Anyone may modify the program for personal use.

\c Permission is granted to copy and distribute modified versions of this
\c manual under the following five pages of conditions...

\c End the Copyleft page and don't use headings on this page.
\clearpage
\pagestyle{headings}

\pagenumbering{roman}
\tableofcontents

\c End the Table of Contents
\clearpage
\pagenumbering{arabic}

\setfilename{plplot.info}
\c Anything before the \setfilename will not appear in the Info file.

\begin{ifinfo}
Put another 5 page copyleft here for the Info file.
\end{ifinfo}

\c The Top node contains the master menu for the Info file.
\c This appears only in the Info file, not the printed manual.

\node Top,       Introduction, (dir),   (dir)

\begin{menu}
This document describes PLplot, the awesome free plotting library for
creating cool scientific plots.

This document is still under construction, so please pardon the mess.

* Introduction::	What PLplot is all about.
* Simple Use::		How to get started.
* Advanced Use::	How to go wild.
* X Windows Drivers::	Xwin, Tk, Tcl-DP, and all the rest.
* Other Drivers::	All non-X drivers.
* API::			C/Fortran API reference.
* Tcl::			How to use PLplot from Tcl.
* Extended WISH::	Building customized WISHes.
* C++::			Using PLplot from C++.
* GUI Development::	Constructing GUI's with PLplot, Tcl and Tk.
* Bibliography::	What you don't want to know about refs.
* Concept Index::	All the stuff you need to know.
\end{menu}

\c %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\part{Introduction}

\input{intro}

\part{Programming}

\input{simple}

\input{advanced}

\input{drivers}

\part{Reference}

\input{api}

\input{tcl}

\part{Advanced Techniques}

\input{wish}

\input{cplus}

\input{gui}

\c %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\part{Closure}

\bibliographystyle{plain}

\node Bibliography, Concept Index, GUI Development, Top

This is the bibliography.

\c \begin{thebibliography}{1}
\c 
\c \bibitem{birdsall}
\c C.~K. Birdsall and A.~B. Langdon.  {\it Plasma Physics via Computer
\c Simulation}.  McGraw Hill, New York. 1985.
\c 
\c \bibitem{hockney}
\c R.~W. Hockney and J.~W. Eastwood.  {\it Computer Simulation Using Particles}.
\c Adam Hilgar, Bristol and New York. 1988.
\c 
\c \bibitem{lebrun}
\c M.~J. LeBrun, M. G. Gray, G. Furnish, T. Tajima, and W. D. Nystrom  {\it
\c The Numerical ``Laboratory'' for Plasma Simulation: A Modern Approach
\c for the Condtruction and Maintenance of a Large Simulation Code}
\c Proceedings of the 14th International Conference on the Numerical
\c Simulation of Plasmas
\c 
\c \end{thebibliography}

\c  \node Concept Index, Top, First Chapter, Top
 \node Concept Index,, Bibliography, Top
\c  \unnumbered{Concept Index}

\twocolumn
\printindex{cp}

\end{document}
